\documentclass[11pt]{article}

\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{amsmath,amssymb,amsthm}
\usepackage{hyperref}
\usepackage[margin=1in]{geometry}
\usepackage{listings}
\usepackage{xcolor}

\lstset{
  basicstyle=\ttfamily\small,
  keywordstyle=\color{blue}\bfseries,
  commentstyle=\color{gray},
  breaklines=true,
  frame=single,
  language=lean
}

\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}[theorem]{Lemma}

\title{A Formal Verification of TCP (RFC~793) in Lean~4}
\author{Samuel Schlesinger}
\date{\today}

\begin{document}

\maketitle

\begin{abstract}
We present a machine-checked formalization of the Transmission Control
Protocol as specified in RFC~793. Using the Lean~4 theorem prover, we model
the TCP state machine, define segment acceptability predicates over modular
32-bit sequence number arithmetic, and prove correctness properties of
connection establishment (three-way handshake), graceful teardown, and global
safety invariants including an unconditional system invariant, send window
bounds, and sequence number monotonicity. We prove that reaching ESTABLISHED
state requires a three-way handshake. The formalization uses a two-endpoint
system model with an explicit network, capturing reordering, loss, and
duplication. All theorems are verified by Lean's kernel with zero uses of
\texttt{sorry}---if the code typechecks, the proofs hold.
\end{abstract}

\tableofcontents

\section{Introduction}
\label{sec:introduction}

The Transmission Control Protocol (TCP) is the foundation of reliable
communication on the Internet. Originally specified by Postel~\cite{rfc793},
TCP provides ordered, reliable, byte-stream delivery over an unreliable
network. Despite its ubiquity, TCP's specification is given in informal prose,
and subtle corner cases---simultaneous open, old duplicate SYN recovery,
half-open connection detection, FIN-with-data---have historically been sources
of implementation bugs.

This work presents a machine-checked formalization of TCP in Lean~4. Our
contributions are:
\begin{enumerate}
  \item A \textbf{modular sequence number arithmetic} library with a circular
    less-than relation and proofs of irreflexivity, decidability, and
    arithmetic interaction properties.
  \item \textbf{Core type definitions} faithfully modeling RFC~793 segments,
    the Transmission Control Block (TCB), and the eleven connection states.
  \item A \textbf{two-endpoint system model} with an explicit network (segment
    multiset) and a \texttt{SystemStep} inductive type capturing delivery,
    user calls, timeouts, loss, and duplication.
  \item The \textbf{full event-processing function} implementing RFC~793
    \S3.9: all user calls (OPEN, SEND, RECEIVE, CLOSE, ABORT, STATUS), segment
    arrival in CLOSED, LISTEN, SYN-SENT, and the 8-check pipeline for
    remaining states, and all three timeout types.
  \item \textbf{Segment and ACK acceptability} predicates with the four-case
    table from RFC~793 \S3.3 and lemmas connecting them to TCB invariants.
  \item \textbf{Handshake correctness}: the three-way handshake reaches
    ESTABLISHED with synchronized sequence numbers, and RST correctly handles
    stale connections.
  \item \textbf{Teardown correctness}: graceful close transitions through the
    correct states, FIN occupies one sequence number, and TIME-WAIT behaves
    correctly.
  \item \textbf{Safety invariants}: no data before ESTABLISHED,
    \texttt{SND.UNA} $\leq$ \texttt{SND.NXT} preservation across handshake
    and close, ACK acceptability preservation, an unconditional system
    invariant with send window bounds, and \texttt{SND.UNA}/\texttt{RCV.NXT}
    monotonicity.
  \item \textbf{Handshake safety property}: ESTABLISHED is only reachable
    from SYN-SENT or SYN-RECEIVED, both of which require exchanging SYN and
    SYN-ACK segments.
\end{enumerate}

All proofs compile with zero uses of \texttt{sorry}. The formalization targets
Lean~4 v4.28.0 without Mathlib.

\section{Sequence Numbers}
\label{sec:seqnum}

TCP sequence numbers are 32-bit unsigned integers with modular arithmetic
(RFC~793 \S3.3). The key challenge is defining a meaningful ordering: since
the space wraps around, the standard total order on natural numbers is
inappropriate.

\begin{definition}[SeqNum]
A \texttt{SeqNum} is a wrapper over \texttt{UInt32}:
\begin{lstlisting}
structure SeqNum where
  val : UInt32
\end{lstlisting}
\end{definition}

\begin{definition}[Circular Less-Than]
$a < b$ iff $b$ is strictly ahead of $a$ within the forward half-space:
\begin{lstlisting}
def lt (a b : SeqNum) : Bool :=
  let d := b.val - a.val
  d != 0 && d < halfSpace  -- halfSpace = 2^31
\end{lstlisting}
This corresponds to $(b - a) \bmod 2^{32} \in (0, 2^{31})$.
\end{definition}

\begin{theorem}[Irreflexivity]
\texttt{SeqNum.lt} is irreflexive: $\forall a,\; \lnot(a < a)$.
This follows because $a - a = 0$, and the predicate requires $d \neq 0$.
Proved as \texttt{SeqNum.lt\_irrefl}.
\end{theorem}

\begin{theorem}[Decidability]
\texttt{SeqNum.lt} is decidable, with a \texttt{Decidable} instance provided
for use in \texttt{if} expressions and \texttt{decide} tactics.
\end{theorem}

We also prove interaction lemmas: \texttt{add\_zero}, \texttt{diff\_self},
and \texttt{le\_refl}.

\section{Core Types}
\label{sec:types}

The \texttt{Tcp.Types} module defines the fundamental structures.

\begin{definition}[Segment]
A TCP segment models the header fields from RFC~793 \S3.1: source/destination
ports, sequence number, acknowledgment number, control bits (URG, ACK, PSH,
RST, SYN, FIN), window size, urgent pointer, and payload data. Wire-format
fields (checksum, data offset, options) are omitted as they concern encoding,
not protocol logic.
\end{definition}

\begin{definition}[SEG.LEN]
Segment length includes data octets plus SYN and FIN contributions to sequence
space (RFC~793 \S3.3): $\texttt{SEG.LEN} = |\texttt{data}| + \texttt{SYN} +
\texttt{FIN}$.
\end{definition}

\begin{definition}[TCB]
The Transmission Control Block stores all send and receive variables from
RFC~793 \S3.2: \texttt{SND.UNA}, \texttt{SND.NXT}, \texttt{SND.WND},
\texttt{SND.UP}, \texttt{SND.WL1}, \texttt{SND.WL2}, \texttt{ISS},
\texttt{RCV.NXT}, \texttt{RCV.WND}, \texttt{RCV.UP}, and \texttt{IRS}.
\end{definition}

\begin{definition}[TcpState]
The eleven TCP states: Closed, Listen, SynSent, SynReceived, Established,
FinWait1, FinWait2, CloseWait, Closing, LastAck, TimeWait.
\end{definition}

\section{State Machine}
\label{sec:state}

\subsection{Endpoint and System Model}

A \texttt{TcpEndpoint} bundles the connection state, TCB, open mode
(active/passive), send queue, receive buffer, reassembly buffer,
retransmission queue, and FIN tracking fields.

The \texttt{System} structure models two communicating endpoints with an
explicit network:
\begin{lstlisting}
structure System where
  endpointA : TcpEndpoint
  endpointB : TcpEndpoint
  network : List Segment
\end{lstlisting}

The network is treated as an unordered multiset: any segment may be delivered
next. This is the standard approach in protocol verification.

\subsection{Event Processing}

Event processing follows RFC~793 \S3.9 faithfully:

\begin{itemize}
  \item \textbf{User calls}: OPEN (active/passive), SEND (segmentize and
    transmit), RECEIVE (deliver buffered data), CLOSE (send FIN), ABORT
    (send RST), STATUS.
  \item \textbf{Segment arrival in CLOSED}: discard data; respond with RST
    unless the incoming segment is RST.
  \item \textbf{Segment arrival in LISTEN}: ignore RST; reject ACK with RST;
    process SYN by sending SYN-ACK and entering SYN-RECEIVED.
  \item \textbf{Segment arrival in SYN-SENT}: the 5-step pipeline (ACK check,
    RST check, security, SYN check, drop).
  \item \textbf{8-check pipeline} for remaining states: (1)~sequence check,
    (2)~RST, (3)~security (skipped), (4)~SYN, (5)~ACK + window update,
    (6)~URG, (7)~segment text, (8)~FIN.
  \item \textbf{Timeouts}: user timeout (abort), retransmission (resend head
    of queue), TIME-WAIT (close).
\end{itemize}

\subsection{SystemStep}

The \texttt{SystemStep} inductive relates system states via atomic transitions:
segment delivery to either endpoint, user calls, timeouts, segment loss, and
segment duplication. A \texttt{SystemTrace} is the reflexive-transitive closure.

\section{Segment Acceptability}
\label{sec:acceptability}

\begin{definition}[Segment Acceptability]
The four-case table from RFC~793 \S3.3:

\begin{center}
\begin{tabular}{ccl}
\textbf{SEG.LEN} & \textbf{RCV.WND} & \textbf{Test} \\
\hline
0 & 0 & $\texttt{SEG.SEQ} = \texttt{RCV.NXT}$ \\
0 & $>0$ & $\texttt{RCV.NXT} \leq \texttt{SEG.SEQ} < \texttt{RCV.NXT} + \texttt{RCV.WND}$ \\
$>0$ & 0 & not acceptable \\
$>0$ & $>0$ & start or end of segment in window \\
\end{tabular}
\end{center}
\end{definition}

\begin{definition}[ACK Acceptability]
Two variants are defined:
\begin{itemize}
  \item \textbf{ESTABLISHED+}: $\texttt{SND.UNA} < \texttt{SEG.ACK} \leq
    \texttt{SND.NXT}$ (strict left inequality---duplicate ACKs excluded).
  \item \textbf{SYN-RECEIVED}: $\texttt{SND.UNA} \leq \texttt{SEG.ACK} \leq
    \texttt{SND.NXT}$ (non-strict---allows $\texttt{SEG.ACK} = \texttt{ISS}$).
\end{itemize}
\end{definition}

\begin{lemma}[Zero-length, zero-window]
A zero-length segment with zero window is acceptable iff
$\texttt{SEG.SEQ} = \texttt{RCV.NXT}$.
Proved as \texttt{segmentAcceptable\_zero\_zero}.
\end{lemma}

\begin{lemma}[Data with zero window]
A data-bearing segment is never acceptable when $\texttt{RCV.WND} = 0$.
Proved as \texttt{segmentAcceptable\_data\_zeroWnd}.
\end{lemma}

\begin{lemma}[ACK preserves invariant]
If an ACK is acceptable (ESTABLISHED-style), advancing \texttt{SND.UNA} to
\texttt{SEG.ACK} preserves $\texttt{SND.UNA} \leq \texttt{SND.NXT}$.
Proved as \texttt{ackAcceptable\_preserves\_invariant}.
\end{lemma}

\section{Connection Establishment}
\label{sec:handshake}

\begin{theorem}[Three-Way Handshake Correctness]
Starting from $A = \text{CLOSED}$ and $B = \text{LISTEN}$, after $A$ performs
active OPEN, the SYN is delivered to $B$, the SYN-ACK is delivered to $A$,
and the ACK is delivered to $B$: both endpoints reach ESTABLISHED with
$A.\texttt{tcb.irs} = B.\texttt{tcb.iss}$ and $B.\texttt{tcb.irs} =
A.\texttt{tcb.iss}$.
Proved symbolically as \texttt{threeWayHandshake\_correct} for arbitrary ISS
values, using a two-pass simplification strategy.
\end{theorem}

\begin{theorem}[SYN Occupies Sequence Space]
After sending a SYN, $\texttt{SND.NXT} = \texttt{ISS} + 1$.
Proved symbolically as \texttt{syn\_occupies\_sequence\_space}.
\end{theorem}

\begin{theorem}[Simultaneous Open]
When both endpoints perform active OPEN and receive the other's SYN in
SYN-SENT, each transitions to SYN-RECEIVED (not ESTABLISHED), as expected
for the first phase of simultaneous open.
Proved as \texttt{simultaneousOpen\_enters\_synReceived}.
\end{theorem}

\begin{theorem}[ESTABLISHED Requires Handshake]
An endpoint can only reach ESTABLISHED by processing a delivered segment while
in SYN-SENT or SYN-RECEIVED state. No user call, timeout, or other state
transition enters ESTABLISHED. Entering SYN-SENT requires an active OPEN;
entering SYN-RECEIVED requires receiving a SYN; entering ESTABLISHED from
SYN-SENT requires a SYN+ACK. These lemmas together prove that ESTABLISHED
is reachable only through the three-way handshake.
Proved as \texttt{established\_requires\_handshake\_step},
\texttt{established\_only\_from\_pre\_established},
\texttt{processUserCall\_not\_established},
and \texttt{processTimeout\_not\_established}.
\end{theorem}

\begin{theorem}[RST Handling]
A valid RST in SYN-RECEIVED with passive origin returns the endpoint to
LISTEN. A valid RST in ESTABLISHED aborts the connection.
Proved as \texttt{rst\_synReceived\_passive\_returns\_to\_listen} and
\texttt{rst\_established\_aborts}.
\end{theorem}

\section{Connection Teardown}
\label{sec:teardown}

\begin{theorem}[CLOSE from ESTABLISHED]
CLOSE from ESTABLISHED transitions to FIN-WAIT-1, emits a FIN segment, and
advances $\texttt{SND.NXT}$ by 1 (FIN occupies one sequence number).
Proved as \texttt{close\_established\_enters\_finWait1} and
\texttt{fin\_occupies\_sequence\_space}.
\end{theorem}

\begin{theorem}[CLOSE-WAIT $\to$ LAST-ACK]
CLOSE from CLOSE-WAIT enters LAST-ACK (not CLOSING). The RFC~793 \S3.9 text
says ``enter CLOSING state'' but Figure~6 (the state diagram) shows
CLOSE-WAIT $\to$ LAST-ACK. We follow Figure~6 as authoritative.
Proved as \texttt{close\_closeWait\_enters\_lastAck}.
\end{theorem}

\begin{theorem}[TIME-WAIT Behavior]
TIME-WAIT timeout transitions to CLOSED. RST in TIME-WAIT causes immediate
transition to CLOSED.
Proved as \texttt{timeWait\_timeout\_closes} and
\texttt{rst\_timeWait\_closes}.
\end{theorem}

\section{Invariants and Safety Properties}
\label{sec:invariants}

\begin{theorem}[No Data Before ESTABLISHED]
In CLOSED, LISTEN, SYN-SENT, and SYN-RECEIVED states, the RECEIVE call never
returns data to the user. Data piggybacked on SYN segments is buffered in the
reassembly buffer and not delivered until the connection reaches ESTABLISHED.
Proved as \texttt{no\_data\_in\_closed}, \texttt{no\_data\_in\_listen},
\texttt{no\_data\_in\_synSent}, \texttt{no\_data\_in\_synReceived}.
\end{theorem}

\begin{theorem}[Sequence Number Monotonicity]
$\texttt{SND.UNA} \leq \texttt{SND.NXT}$ is preserved by every processing
function. Per-function preservation theorems are proved symbolically:
\texttt{processOpen\_preserves}, \texttt{processSegment\_preserves},
\texttt{processClose\_preserves}, \texttt{processSend\_preserves},
\texttt{processTimeout\_preserves}, etc.
\end{theorem}

\begin{theorem}[ACK Acceptability Preserves Monotonicity]
If an ACK is acceptable (ESTABLISHED-style), advancing $\texttt{SND.UNA}$ to
$\texttt{SEG.ACK}$ maintains $\texttt{SND.UNA} \leq \texttt{SND.NXT}$.
This is the key lemma for proving monotonicity inductively over system traces.
Proved as \texttt{ackAdvanceTcb\_preserves}.
\end{theorem}

\begin{theorem}[Window Invariant]
In-flight bytes (\texttt{SND.NXT} $-$ \texttt{SND.UNA}) remain bounded within
the forward half-space with room for future advances. Pre-FIN states maintain
a bound of $+2 < 2^{31}$; post-FIN states (where no more data will be sent)
maintain $+1 < 2^{31}$. This is proved preserved by every processing function.
Proved as \texttt{windowInvariant} with per-function preservation theorems.
\end{theorem}

\begin{theorem}[SystemStep Invariant (Unconditional)]
The combined system invariant---$\texttt{SND.UNA} \leq \texttt{SND.NXT}$,
the window bound, and the ISS structural invariant for SYN-SENT---is preserved
by every \texttt{SystemStep} transition \emph{unconditionally}, with no
blanket hypotheses. The window invariant provides the half-space room needed
for CLOSE's FIN, eliminating the need for external send-window assumptions.
Proved as \texttt{systemStep\_preserves\_invariant}.
\end{theorem}

\begin{theorem}[SND.UNA Monotonicity]
For any \texttt{SystemStep}, each endpoint's \texttt{SND.UNA} is monotonically
non-decreasing (modulo connection teardown to CLOSED or fresh connection
initialization from CLOSED/LISTEN).
Proved as \texttt{systemStep\_sndUna\_monotone}.
\end{theorem}

\begin{theorem}[RCV.NXT Monotonicity]
For any \texttt{SystemStep} with well-formed segments (data length
$< 2^{31} - 1$), each endpoint's \texttt{RCV.NXT} is monotonically
non-decreasing (modulo connection teardown or initialization from
CLOSED/LISTEN/SYN-SENT).
Proved as \texttt{systemStep\_rcvNxt\_monotone}.
\end{theorem}

\section{Conclusion}
\label{sec:conclusion}

We have presented a machine-checked formalization of TCP (RFC~793) in Lean~4,
covering modular sequence number arithmetic, the full event-processing state
machine, segment acceptability, connection establishment, graceful teardown,
and safety invariants. All proofs compile with zero \texttt{sorry}.

\subsection{Design Decisions}

Key modeling choices include: representing sequence numbers as \texttt{UInt32}
wrappers with a circular half-space comparison; using a two-endpoint system
model with an explicit segment network; implementing event processing as pure
functions with nondeterminism confined to the \texttt{SystemStep} inductive;
and omitting security/precedence checks (RFC~793 \S3.6) as orthogonal
IP-layer concerns.

\subsection{Limitations and Known Deviations}

This formalization intentionally excludes several RFC~793 features:
security/compartment/precedence (\S3.6), checksum computation, MSS option
negotiation, ISN clock-based generation (\S3.3), RTO calculation (\S3.7),
and IP-layer interactions.

Known deviations from RFC~793:
\begin{itemize}
  \item \textbf{CLOSE-WAIT $+$ CLOSE}: the \S3.9 text says ``enter CLOSING
    state'' but Figure~6 (the state diagram) shows CLOSE-WAIT $\to$ LAST-ACK.
    We follow Figure~6 as authoritative. This is a known errata in RFC~793.
  \item \textbf{Urgent pointer interpretation}: RFC~793 \S3.9 says
    \texttt{RCV.UP <- max(RCV.UP, SEG.UP)}, where \texttt{SEG.UP} is defined
    as a positive offset from \texttt{SEG.SEQ}. We interpret this as
    \texttt{RCV.UP <- max(RCV.UP, RCV.NXT + SEG.UP)}, converting the offset
    to a sequence number. This is consistent with the idealized segment
    assumption (\texttt{SEG.SEQ = RCV.NXT}) and is a known RFC ambiguity.
  \item \textbf{No timer state}: the formalization does not model the
    retransmission timer, 2$\times$MSL timer, or user timeout timer as
    explicit state. Timeouts are modeled as nondeterministic events via
    \texttt{SystemStep}.
  \item \textbf{No segment reassembly}: the model assumes in-order segment
    delivery for data processing. Out-of-order segments are buffered in
    \texttt{reassemblyBuf} but never reassembled; the \texttt{pipelineText}
    function appends data directly to \texttt{recvBuffer} and advances
    \texttt{RCV.NXT} by the segment's data length. A full reassembly
    implementation would require tracking sequence number ranges and merging
    overlapping segments, which is orthogonal to the protocol state machine
    properties proved here.
\end{itemize}

\subsection{Future Work}

Natural extensions include: congestion control (RFC~5681); window scaling and
timestamps (RFC~7323); modeling of the full simultaneous open handshake
completion through SYN-RECEIVED to ESTABLISHED; segment reassembly for
out-of-order delivery; and data delivery correctness (proving that the
received byte stream matches the sent byte stream in order).

\bibliographystyle{plain}
\bibliography{references}

\end{document}
